---
title: "Metal Pollution in the US"
author: "Guang Yang"
date: "December 8, 2016"
output: ioslides_presentation
---

## Introduction

This ioslides presentation depicts levels of pollutions of metal dusts, by states in the US.

There are 13 types of metal dusts: `Barium (Ba)`, `Beryllium (Be)`, `Cadmium (Cd)`, `Calcium (Ca)`, `Chromium (Cr)`, `Copper (Cu)`, `Iron (Fe)`, `Lead (Pb)`, `Manganese (Mn)`, `Molybdenum (Mo)`, `Nickel (Ni)`, `Vanadium (V)`, `Zinc (Zn)`.

Data on map are averaged measurements from all monitors in the same state, and with unit of `mg/m^3`.

Processed R data object (`DataGalleryMetal.rda`) contains all data from 1990 to 2016; only the data of 2015 is shown in this presentation (data of 2016 is incomplete), with `Cadmium (Cd)` and `Lead (Pb)` maps in 1990 is provided for comparison, right after their maps of 2015.

## Sources of Data

This document originates some cleaned data from the annual pollution data from [EPA Air-Data webpage](https://www.epa.gov/outdoor-air-quality-data). For more information of [data downloading](https://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/download_files.html) or [data documentation](https://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/FileFormats.html#_annual_summary_files), please follow the links to the EPA websites.

## Access to Codes

The [DataGalleryMetal.rda](https://github.com/imreyes/datasciencecoursera/blob/master/DevelopingDataProduct/Week3_ProgrammingAssignment/DataGalleryMetal.rda) stores the list named `data.gallery.metal` which contains the necessary data chunks for mapping. Codes to generate the list is stored in [GasPollutionMetal_DataProcessor.R](https://github.com/imreyes/datasciencecoursera/blob/master/DevelopingDataProduct/Week3_ProgrammingAssignment/GasPollutionMetal_DataProcessor.R) file. 

```{r load data, echo=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(plotly))

# Fetch the data object from the web.
if(!file.exists('DataGalleryMetal.rda')) {
        Url <- 'https://github.com/imreyes/datasciencecoursera/raw/master/DevelopingDataProduct/Week3_ProgrammingAssignment/DataGalleryMetal.rda'
        download.file(Url, destfile = 'DataGalleryMetal.rda')
}
load('DataGalleryMetal.rda')

# Make white border between states.
border <- list(color = toRGB('black'), lwd = 2)
# Specify Map options.
opt <- list(scope = 'usa', projection = list(type = 'albers usa'),
            showlakes = TRUE, lakecolor = toRGB('white'))

```

## Cadmium Pollution in 2015

Note: not all states have reported data.

```{r Cd_2015, echo=FALSE, warning=FALSE}
# Read-in data. 
# rows are 4 variables of 13 chemicals; columns are year minus 1989.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*2+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Cadmium Pollution in 2015', geo = opt)
```

## Cadmium Pollution in 1990

Note: not all states have reported data.

```{r Cd_1990, echo=FALSE, warning=FALSE}
# Read-in data. 
# rows are 4 variables of 13 chemicals; columns are year minus 1989.
dgm <- as.data.frame(t(data.gallery.metal)[1, 4*2+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Cadmium Pollution in 1990', geo = opt)
```

## Lead Pollution in 2015

Note: not all states have reported data.

```{r Pb_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*7+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Lead Pollution in 2015', geo = opt)
```

## Lead Pollution in 1990

Note: not all states have reported data.

```{r Pb_1990, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[1, 4*7+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Lead Pollution in 1990', geo = opt)
```

## Barium Pollution in 2015

Note: not all states have reported data.

```{r Ba_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*0+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Barium Pollution in 2015', geo = opt)
```

## Beryllium Pollution in 2015

Note: not all states have reported data.

```{r Be_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*1+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Barium Pollution in 2015', geo = opt)
```

## Calcium Pollution in 2015

Note: not all states have reported data.

```{r Ca_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*3+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Calcium Pollution in 2015', geo = opt)
```

## Chromium Pollution in 2015

Note: not all states have reported data.

```{r Cr_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*4+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Chromium Pollution in 2015', geo = opt)
```

## Copper Pollution in 2015

Note: not all states have reported data.

```{r Cu_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*5+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Copper Pollution in 2015', geo = opt)
```

## Iron Pollution in 2015

Note: not all states have reported data.

```{r Fe_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*6+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Iron Pollution in 2015', geo = opt)
```

## Manganese Pollution in 2015

Note: not all states have reported data.

```{r Mn_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*8+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Manganese Pollution in 2015', geo = opt)
```

## Molybdenum Pollution in 2015

Note: not all states have reported data.

```{r Mo_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*9+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Molybdenum Pollution in 2015', geo = opt)
```

## Vanadium Pollution in 2015

Note: not all states have reported data.

```{r Ni_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*11+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Vanadium Pollution in 2015', geo = opt)
```

## Zinc Pollution in 2015

Note: not all states have reported data.

```{r Zn_2015, echo=FALSE, warning=FALSE}
# Read-in data.
# rows are 4 variables of 13 chemicals; columns are year minus 1990.
dgm <- as.data.frame(t(data.gallery.metal)[26, 4*12+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Zinc Pollution in 2015', geo = opt)
```

## Conclusion

* Lead and cadmium pollution is greately reduced (over 10-fold) over the 26 years; the cadmium pollution is significant in Missouri.

* Texas, Missouri, Michigan suffered most from pollution of metal dusts, in general. Minnesota and Kentucky also bore high-level pollution of specific metal dust, comparing to other states.

# Thank you!