---
title: "Statistical Inference Week4 Programming Assignment (1)"
author: "Guang Yang"
date: "November 17, 2016"
output:
  html_document: default
  pdf_document: default
---

# Statistical Inference Week4 Programming Assignment (1)


## Overview

This report addresses the questions from [Week 4 Assignment](https://www.coursera.org/learn/statistical-inference/peer/3k8j5/statistical-inference-course-project) of __Statistical Inference__, the Course #6 of the Data Science Specialization series, offered by [Coursera.org](https://www.coursera.org/). The report is mainly consist of 2 parts, aiming to discuss the results of the two topics:

* __Testing CLT with simulation from exponential distribution__;

* Performing inferential analysis on the ToothGrowth dataset.


## Case Study 1: Simulation of CLT Based on Exponential Equation

In this project we will investigate the exponential distribution in R and compare it with the Central Limit Theorem (CLT). The exponential distribution can be simulated in R with `rexp(n, lambda)` where `lambda` is the rate parameter. The mean of exponential distribution is `1/lambda` and the standard deviation is also `1/lambda`. Set `lambda = 0.2` for all of the simulations. We will investigate the distribution of averages of 40 exponentials. Note that we will do __1000__ simulations.

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponentials.

```{r ExpSim40, echo = FALSE}
n <- 40                 # Sample Size
times <- 1000           # Repeat Time
lambda <- 0.2           # Rate Constant

# Use set.seed() for reproducible purpose.
set.seed(7)
means <- replicate(times, mean(rexp(n, lambda)))

# Prepare numeric sequence to plot normal curve.
LIM <- c(floor(min(means)), ceiling(max(means)))
x <- seq(floor(min(means)), ceiling(max(means)), length = 1000)
nd <- dnorm(x, mean = 1/lambda, sd = 1/(lambda*sqrt(n)))

# Make plot.
library(ggplot2)
ggplot(as.data.frame(means), aes(x = means)) +
        geom_histogram(bins = 30, col = 'black', fill = 'blue') +
        lims(x = LIM) +
        # Theoretical Mean
        geom_vline(xintercept = 1/lambda, col = 'red', lwd = 2) +
        # Sample Mean
        geom_vline(xintercept = mean(means), col = 'yellow', lwd = 2) +
        # Theoretical SD
        geom_vline(xintercept = 1/lambda + c(-1, 1) * 1/(lambda*sqrt(n)), col = 'red') +
        # Sample SD
        geom_vline(xintercept = mean(means) + c(-1, 1) * sd(means), col = 'yellow') +
        # Gaussian PDF
        geom_line(aes(x = x, y = nd*times/5), col = 'green', lwd = 2) +
        labs(x = 'Means', y = 'Counts',
             title = 'Distribution of 1000 40-Mean Samples from Exponential Distribution')

svar <- round(var(means), 3)
tvar <- round(1/lambda^2/n, 3)
print(paste0('Theoretical variance: ', tvar))
print(paste0('Sample variance: ', svar))
```

From the above plot we firstly see the __distribution of the sample means__ (blue); the top of which stands somewhere between 4 and 5. The above plot:

1. Shows the __sample mean__, `mean(means)` __(yellow)__  and compare it to the __theoretical mean__, `1/lambda` __(red)__ of the distribution.

2. Shows how variable the sample is and compare it to the theoretical variance of the distribution.

        + By variance: the variance of sample means is `r svar`, and that of theoretical means is `r tvar`.
        + By standard deviation: the standard deviations in the plot are marked in orange (sample) and red (theoretical).

3. Shows that the distribution is approximately normal by overlaying the __gaussian PDF__ (`mean = 1/lambda, sd = 1/(lambda*sqrt(n))`, green curve) with the histogram.


Just to make a comparison, we then quickly plot another set of 1000 samples from distribution. _Note this time there's only 1 sample each, instead of mean of 40_. Let's take a look at it.
```{r ExpSim1, echo = FALSE}
times <- 1000           # Repeat Time
lambda <- 0.2           # Rate Constant

# Use set.seed() for reproducible purpose.
set.seed(7)
means2 <- replicate(times, rexp(1, lambda))

# Prepare numeric sequence to plot normal curve.
LIM2 <- c(floor(min(means2)), ceiling(max(means2)))
x2 <- seq(floor(min(means2)), ceiling(max(means2)), length = 1000)
nd2 <- dnorm(x2, mean = 1/lambda, sd = 1/lambda)

# Make plot.
library(ggplot2)
ggplot(as.data.frame(means2), aes(x = means2)) +
        geom_histogram(bins = 30, col = 'black', fill = 'blue') +
        lims(x = LIM2) +
        # Theoretical Mean
        geom_vline(xintercept = 1/lambda, col = 'red', lwd = 2) +
        # Sample Mean
        geom_vline(xintercept = mean(means2), col = 'yellow', lwd = 2) +
        # Theoretical SD
        geom_vline(xintercept = 1/lambda + c(-1, 1) * 1/(lambda), col = 'red') +
        # Sample SD
        geom_vline(xintercept = mean(means2) + c(-1, 1) * sd(means2), col = 'yellow') +
        # Gaussian PDF
        geom_line(aes(x = x2, y = nd2*times*1.5), col = 'green', lwd = 2) +
        labs(x = 'Means', y = 'Counts',
             title = 'Distribution of 1000 samples from Exponential Distribution')

svar2 <- round(var(means2), 3)
tvar2 <- round(1/lambda^2, 3)
print(paste0('Theoretical variance: ', tvar2))
print(paste0('Sample variance: ', svar2))
```

Now as we see in the above, the distribution is way much more like exponential, rather than normal, even if the variances are similar (`r tvar2` of theoretical and `r svar2` of sample). In comparison, the 40-mean distribution is very normal-like.

















## Appendix: Codes

### Codes for Plotting "Distribution of 1000 40-Mean Samples from Exponential Distribution"

```{r ExpSim40_App, eval = FALSE}
n <- 40                 # Sample Size
times <- 1000           # Repeat Time
lambda <- 0.2           # Rate Constant

# Use set.seed() for reproducible purpose.
set.seed(7)
means <- replicate(times, mean(rexp(n, lambda)))

# Prepare numeric sequence to plot normal curve.
LIM <- c(floor(min(means)), ceiling(max(means)))
x <- seq(floor(min(means)), ceiling(max(means)), length = 1000)
nd <- dnorm(x, mean = 1/lambda, sd = 1/(lambda*sqrt(n)))

# Make plot.
library(ggplot2)
ggplot(as.data.frame(means), aes(x = means)) +
        geom_histogram(bins = 30, col = 'black', fill = 'blue') +
        lims(x = LIM) +
        # Theoretical Mean
        geom_vline(xintercept = 1/lambda, col = 'red', lwd = 2) +
        # Sample Mean
        geom_vline(xintercept = mean(means), col = 'yellow', lwd = 2) +
        # Theoretical SD
        geom_vline(xintercept = 1/lambda + c(-1, 1) * 1/(lambda*sqrt(n)), col = 'red') +
        # Sample SD
        geom_vline(xintercept = mean(means) + c(-1, 1) * sd(means), col = 'yellow') +
        # Gaussian PDF
        geom_line(aes(x = x, y = nd*times/5), col = 'green', lwd = 2) +
        labs(x = 'Means', y = 'Counts',
             title = 'Distribution of 1000 40-Mean Samples from Exponential Distribution')

svar <- round(var(means), 3)
tvar <- round(1/lambda^2/n, 3)
print(paste0('Theoretical variance: ', tvar))
print(paste0('Sample variance: ', svar))
```

### Codes for Plotting "Distribution of 1000 samples from Exponential Distribution"

```{r ExpSim1_App, eval = FALSE}
times <- 1000           # Repeat Time
lambda <- 0.2           # Rate Constant

# Use set.seed() for reproducible purpose.
set.seed(7)
means2 <- replicate(times, rexp(1, lambda))

# Prepare numeric sequence to plot normal curve.
LIM2 <- c(floor(min(means2)), ceiling(max(means2)))
x2 <- seq(floor(min(means2)), ceiling(max(means2)), length = 1000)
nd2 <- dnorm(x2, mean = 1/lambda, sd = 1/lambda)

# Make plot.
library(ggplot2)
ggplot(as.data.frame(means2), aes(x = means2)) +
        geom_histogram(bins = 30, col = 'black', fill = 'blue') +
        lims(x = LIM2) +
        # Theoretical Mean
        geom_vline(xintercept = 1/lambda, col = 'red', lwd = 2) +
        # Sample Mean
        geom_vline(xintercept = mean(means2), col = 'yellow', lwd = 2) +
        # Theoretical SD
        geom_vline(xintercept = 1/lambda + c(-1, 1) * 1/(lambda), col = 'red') +
        # Sample SD
        geom_vline(xintercept = mean(means2) + c(-1, 1) * sd(means2), col = 'yellow') +
        # Gaussian PDF
        geom_line(aes(x = x2, y = nd2*times*1.5), col = 'green', lwd = 2) +
        labs(x = 'Means', y = 'Counts',
             title = 'Distribution of 1000 samples from Exponential Distribution')

svar2 <- round(var(means2), 3)
tvar2 <- round(1/lambda^2, 3)
print(paste0('Theoretical variance: ', tvar2))
print(paste0('Sample variance: ', svar2))
```