---
title: "Statistical Inference Week4 Programming Assignment (2)"
author: "Guang Yang"
date: "November 17, 2016"
output:
  html_document: default
  pdf_document: default
---

# Statistical Inference Week4 Programming Assignment (2)

## Overview

This report addresses the questions from [Week 4 Assignment](https://www.coursera.org/learn/statistical-inference/peer/3k8j5/statistical-inference-course-project) of __Statistical Inference__, the Course #6 of the Data Science Specialization series, offered by [Coursera.org](https://www.coursera.org/). The report is mainly consist of 2 parts, aiming to discuss the results of the two topics:

* Testing CLT with simulation from exponential distribution;

* __Performing inferential analysis on the ToothGrowth dataset__.


## Case Study 2: 

In this case we will take a look at the `ToothGrowth` dataset from `datasets` package.

First let's load and take a quick look at the dataset.

```{r TG_getdata, echo = FALSE}
# Load dataset
library(datasets)
data(ToothGrowth)

# Look at the dataset
dim(ToothGrowth)
summary(ToothGrowth)
str(ToothGrowth)
boxplot(ToothGrowth$len ~ ToothGrowth$supp, xlab = 'Supplement Type',
        ylab = 'Tooth Length',
        main = 'Tooth Lengths with Different Supplement Types')
```

```{r TG_sep_plot, echo = FALSE}
boxplot(ToothGrowth$len ~ ToothGrowth$dose, xlab = 'Supplement Dose',
        ylab = 'Tooth Length',
        main = 'Tooth Lengths with Different Supplement Doses')
```

So from the summary we know that the dataset has 3 variables: 

* len  - tooth length: numeric value indicating tooth growth;

* supp - supplement type: 2-component factor (`VC`, `OJ`), 30 entries each;

* dose - dose in mg/day: 3-component numeric factor (`0.5`, `1`, `2`), 20 entries each.

Now we need to perform inferential analysis over the topics:

1. Is vitamin supplement type related to tooth growth?

2. Is vitamin supplement dose related to tooth growth?

For question 1, we'll establish H0: there's no difference between the 2 groups of supplements. Then we'll use t-test to find out p-value, and compare to the cutoff alpha = 0.05 (two sides).

```{r ToothGrowth, echo = FALSE}
# Is vitamin supplement type related to tooth growth?
ttest <- t.test(len ~ supp, data = ToothGrowth)
ttest
```

As the p-value (`r ttest$p.val`) is greater, and confidence interval (`r round(ttest$conf[1], 2)` ~ `r round(ttest$conf[2], 2)`) passes 0, we therefore have to retain the H0: __difference in supplement types isn't relevant to tooth growth__.

And for question 2, we need to do the same thing, but one pair each time; hence there are 3 null hypotheses to be checked. We'll show the summary of only 1 here.

```{r ToothGrowth2, echo = FALSE}
# Is vitamin supplement dose related to tooth growth?
# Here we show comparison between 0.5 and 1.
dose1 <- ToothGrowth[ToothGrowth$dose == 0.5, 1]
dose2 <- ToothGrowth[ToothGrowth$dose == 1, 1]
ttest1 <- t.test(dose1, dose2)
ttest1

# p-values for the other pairs
dose3 <- ToothGrowth[ToothGrowth$dose == 2, 1]
ttest2 <- t.test(dose1, dose3)
ttest3 <- t.test(dose2, dose3)
```

As we see, the p-value between dose of 0.5 mg/d and 1 mg/d is `r ttest1$p.val`, way smaller than 0.05; so do the comparisons between 0.5 mg/d and 2 mg/d (`r ttest2$p.val`), and 1 mg/d and 2 mg/d (`r ttest3$p.val`). Hence we can safely reject the H0, and confirm that __the dose does have a relationship with tooth growth__.





## Appendix: Codes

### Codes for Data Loading, Summary and Plotting

```{r TG_getdata_App, eval = FALSE}
# Load dataset
library(datasets)
data(ToothGrowth)

# Look at the dataset
dim(ToothGrowth)
summary(ToothGrowth)
str(ToothGrowth)
boxplot(ToothGrowth$len ~ ToothGrowth$supp, xlab = 'Supplement Type',
        ylab = 'Tooth Length',
        main = 'Tooth Lengths with Different Supplement Types')
boxplot(ToothGrowth$len ~ ToothGrowth$dose, xlab = 'Supplement Dose',
        ylab = 'Tooth Length',
        main = 'Tooth Lengths with Different Supplement Doses')
```

### Codes for T-Test of Supplement Types

```{r ToothGrowth_App, eval = FALSE}
# Is vitamin supplement type related to tooth growth?
ttest <- t.test(len ~ supp, data = ToothGrowth)
ttest
```

### Codes for T-Tests of Supplement Doses

```{r ToothGrowth2_App, eval = FALSE}
# Is vitamin supplement dose related to tooth growth?
# Here we show comparison between 0.5 and 1.
dose1 <- ToothGrowth[ToothGrowth$dose == 0.5, 1]
dose2 <- ToothGrowth[ToothGrowth$dose == 1, 1]
ttest1 <- t.test(dose1, dose2)
ttest1

# p-values for the other pairs
dose3 <- ToothGrowth[ToothGrowth$dose == 2, 1]
ttest2 <- t.test(dose1, dose3)
ttest3 <- t.test(dose2, dose3)
```