---
title: "Gas and Metal Dust Pollution in the US (1990 - 2016)"
author: "Guang Yang"
date: "December 9, 2016"
output: ioslides_presentation
---

## Introduction

This ioslides presentation serves to further document and show the [shiny web application of the same name](https://imreyes.shinyapps.io/GasMetalPollution/).

The web application visualizes pollutions of 3 types of gases and 13 types of metal dusts, by states in the US.

The 13 types of metal dusts are: `Barium (Ba)`, `Beryllium (Be)`, `Cadmium (Cd)`, `Calcium (Ca)`, `Chromium (Cr)`, `Copper (Cu)`, `Iron (Fe)`, `Lead (Pb)`, `Manganese (Mn)`, `Molybdenum (Mo)`, `Nickel (Ni)`, `Vanadium (V)`, `Zinc (Zn)`.

The 3 types of gases are: `Sulfur dioxide`, `Carbon monoxide`, `Ozone`.

## Sources of Data

This document originates some cleaned data from the annual pollution data from [EPA Air-Data webpage](https://www.epa.gov/outdoor-air-quality-data). For more information of [data downloading](https://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/download_files.html) or [data documentation](https://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/FileFormats.html#_annual_summary_files), please follow the links to the EPA websites.

The [DataGalleryGasMetal.rda](https://github.com/imreyes/datasciencecoursera/blob/master/DevelopingDataProduct/Week4_ProgrammingAssignment/DataGalleryGasMetal.rda) stores the list named `data.gallery.metal` which contains the necessary data chunks for mapping. Codes to generate the list is stored in [GasMetalPollution_DataProcessor.R](https://github.com/imreyes/datasciencecoursera/blob/master/DevelopingDataProduct/Week4_ProgrammingAssignment/GasMetalPollution_DataProcessor.R) file. 

Hovering on the map will show relevant data, including `state name`, `pollution level (Mean)` and `Number of monitor sites in the state (# of Sites)`.

```{r load data, echo=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(plotly))

# Fetch the data object from the web.
if(!file.exists('DataGalleryGasMetal.rda')) {
        Url <- 'https://github.com/imreyes/datasciencecoursera/raw/master/DevelopingDataProduct/Week4_ProgrammingAssignment/DataGalleryGasMetal.rda'
        download.file(Url, destfile = 'DataGalleryGasMetal.rda')
}
load('DataGalleryGasMetal.rda')

# Make white border between states.
border <- list(color = toRGB('black'), lwd = 2)
# Specify Map options.
opt <- list(scope = 'usa', projection = list(type = 'albers usa'),
            showlakes = TRUE, lakecolor = toRGB('white'))

```

## Cadmium Pollution in 2015

Note: not all states have reported data.

```{r Cd_2015, echo=FALSE, warning=FALSE}
# Read-in data. 
# rows are 4 variables of 13 chemicals; columns are year minus 1989.
dgm <- as.data.frame(t(data.gallery.all)[26, 4*2+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Cadmium Pollution in 2015 (Unit: mg/m3)', geo = opt)
```

## Example: Ozone Pollution in 1990

Note: not all states have reported data.

```{r Cd_1990, echo=FALSE, warning=FALSE}
# Read-in data. 
# rows are 4 variables of 13 chemicals; columns are year minus 1989.
dgm <- as.data.frame(t(data.gallery.all)[1, 4*15+1:4])
names(dgm) <- c('State', 'Mean', 'Num.Monitors', 'hover')
# Plot map
plot_ly(z = dgm$Mean, text = dgm$hover, locations = dgm$State,
        type = 'choropleth', locationmode = 'USA-states',
        color = dgm$Mean, colors = 'Blues', marker = list(line = border)) %>%
        layout(title = 'Ozone Pollution in 1990 (ppm)', geo = opt)
```